---
title: "Visualising relationships"
---

In this tutorial we will continue to learn about visualisation as a tool for exploratory data analysis.
We will look at ways of visualising the <mark>relationship</mark> between two or more variables using bar and column plots, scatterplots, additional aesthetics and <mark>facets</mark>.

## Objectives

By the end of this tutorial you should:

* Be able to generate questions about the relationship between two or more variables
* Know how to produce bar plots, multiple density plots, stacked bar plots, and scatter plots in R using ggplot2
* Be able to refine plots for communication and export them from R

## Prerequisites

* Edward Tufte, *[The Visual Display Of Quantitative Information](https://archive.org/details/the-visual-display-of-quantitative-information-2ed-by-tufte-edward-r.-z-lib.org)* (2nd edition), pp. 91–138:
  * Chapter 4, "Data–Ink and Graphical Redesign"
  * Chapter 5, "Chartjunk: Vibrations, Grids, and Ducks"
  * Chapter 6, "Data–Ink Maximization and Graphical Design"

<!-- 
## Generating questions about relationships

Last week we looked at using visualisation to answer questions about the variation of a variable (its <mark>distribution</mark>).
Although essential for describing and understanding the nature of your dataset, questions about a single variable have a fundamentally limited explanatory value.

This week we will start looking at the <mark>covariation</mark> between two (or more) variables – in plain terms, the relationship between them.
With this we can start to gain insights into possible <mark>causality</mark>

* Questions: correlation and causation
  * Some examples
  * Correlation ≠ causation
  * When do you need models

## Lithic assemblages from Islay

For this tutorial we will use the `islay_lithics` dataset.
Load it with:

```{r}
library(islay)
data(islay_lithics)
```

We can use the `head()` function to get a quick preview of the data frame:

```{r}
head(islay_lithics)
```

Because this is an in-built dataset of the package.

As with the last dataset, it will be useful to turn the `period` column into a <mark>factor</mark> now, so that it will automatically be ordered in our subsequent plots:

```{r}
periods <- c("Mesolithic", "Mesolithic & Later Prehistoric", "Later Prehistoric")
islay_lithics$period <- factor(islay_lithics$period, periods)
```

### Exercises

1. What units is the area of the site given in in the `islay_lithics` dataset?

## Visualising relationships

* Numerical and categorical: box plots or multiple density
* Two categorical: stacked bar plot
* Two numerical: scatter plots
* Three or more variables: additional aesthetics, facets

## Exporting plots

* Think about what Tufte says

-->
